---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/khufkens/MODISTools.svg)](https://travis-ci.org/khufkens/MODISTools) [![codecov](https://codecov.io/gh/khufkens/MODISTools/branch/master/graph/badge.svg)](https://codecov.io/gh/khufkens/MODISTools) ![Status](https://www.r-pkg.org/badges/version/MODISTools) ![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/MODISTools)

# MODISTools

Programmatic interface to the ['MODIS Land Products Subsets' web services](https://modis.ornl.gov/data/modis_webservice.html). Allows for easy downloads of ['MODIS'](http://modis.gsfc.nasa.gov/) time series directly to your R workspace or your computer. When using the package please cite the manuscript as referenced below.

## Installation

### stable release

To install the current stable release use a CRAN repository:

```{r eval = FALSE}
install.packages("MODISTools")
library("MODISTools")
```

### development release

To install the development releases of the package run the following commands:

```{r eval = FALSE}
if(!require(devtools)){install.package("devtools")}
devtools::install_github("khufkens/MODISTools")
library("MODISTools")
```

Vignettes are not rendered by default, if you want to include additional documentation please use:

```{r eval = FALSE}
if(!require(devtools)){install.package("devtools")}
devtools::install_github("khufkens/MODISTools", build_vignettes = TRUE)
library("MODISTools")
```

## Use

### Downloading MODIS time series

To extract a time series of modis data for a given location and its direct environment use the mt_subset() function.

<details><summary>detailed parameter description (click to expand)</summary>
<p>

Parameter     | Description                      
------------- | ------------------------------ 	
product	      | a MODIS product
band	      | a MODIS product band (if NULL all bands are downloaded)
lat           | latitude of the site
lon           | longitude of the site
start      | start year of the time series (data start in 1980)
end        | end year of the time series (current year - 2 years, use force = TRUE to override)
internal      | logical, TRUE or FALSE, if true data is imported into R workspace otherwise it is downloaded into the current working directory
out_dir | path where to store the data when not used internally, defaults to tempdir()
km_lr | force "out of temporal range" downloads (integer)
km_ab | suppress the verbose output (integer)
site_name | a site identifier
site_id | a site_id for predefined locations (not required)
progress | logical, TRUE or FALSE (show download progress)

</p>
</details>

```{r eval = TRUE}
# load the library
library(MODISTools)

# download data
subset <- mt_subset(product = "MOD11A2",
                    lat = 40,
                    lon = -110,
                    band = "LST_Day_1km",
                    start = "2004-01-01",
                    end = "2004-02-01",
                    km_lr = 1,
                    km_ab = 1,
                    site_name = "testsite",
                    internal = TRUE,
                    progress = FALSE)
print(str(subset))
```

The output format is a *tidy* data frame, as shown above. When witten to a csv with the parameter `internal = FALSE` this will result in a flat file on disk.

Note that when a a region is defined using km_lr and km_ab multiple pixels might be returned. These are indexed using the `pixel` column in the data frame containing the time series data. The remote sensing values are listed in the `value` column. When no band is specified all bands of a given product are returned, be mindful of the fact that different bands might require different multipliers to represent their true values. To list all available products, bands for particular products and temporal coverage see function descriptions below.

### Batch downloading MODIS time series

When a large selection of locations is needed you might benefit from using the batch download function `mt_batch_subset()`, which provides a wrapper around the `mt_subset()` function in order to speed up large download batches. This function has a similar syntax to `mt_subset()` but requires a data frame defining site names (site_name) and locations (lat / lon) (or a comma delimited file with the same structure) to specify a list of download locations.

Below an example is provided on how to batch download data for a data frame of given site names and locations (lat / lon).

```{r eval = TRUE}
# create data frame with a site_name, lat and lon column
# holding the respective names of sites and their location
df <- data.frame("site_name" = paste("test",1:2))
df$lat <- 40
df$lon <- -110
  
# test batch download
subsets <- mt_batch_subset(df = df,
                     product = "MOD11A2",
                     band = "LST_Day_1km",
                     internal = TRUE,
                     start = "2004-01-01",
                     end = "2004-02-01")

print(str(subsets))
```

### Listing products

To list all available products use the mt_products() function.

```{r eval = TRUE}
products <- mt_products()
head(products)
```

### Listing bands

To list all available bands for a given product use the mt_bands() function.

```{r eval = TRUE}
bands <- mt_bands(product = "MOD11A2")
head(bands)
```

### listing dates

To list all available dates (temporal coverage) for a given product and location use the mt_dates() function.

```{r eval = TRUE}
dates <- mt_dates(product = "MOD11A2", lat = 42, lon = -110)
head(dates)
```

## References

Tuck et al. (2014). [MODISTools - downloading and processing MODIS remotely sensed data in R Ecology & Evolution](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.1273), 4(24), 4658 - 4668.

## Acknowledgements

Original development was supported by the UK Natural Environment Research Council (NERC; grants NE/K500811/1 and NE/J011193/1), and the Hans Rausing Scholarship. Refactoring was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE).
