---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/khufkens/MODISTools.svg)](https://travis-ci.org/khufkens/MODISTools) [![codecov](https://codecov.io/gh/khufkens/MODISTools/branch/master/graph/badge.svg)](https://codecov.io/gh/khufkens/MODISTools) 

# MODISTools

Programmatic interface to the ['MODIS Land Products Subsets' web services](https://modis.ornl.gov/data/modis_webservice.html). Allows for easy downloads of 'MODIS' time series directly to your R workspace or your computer.

## Installation

### development release

To install the development releases of the package run the following commands:

```{r eval = FALSE}
if(!require(devtools)){install.package("devtools")}
devtools::install_github("khufkens/MODISTools")
library("MODISTools")
```

Vignettes are not rendered by default, if you want to include additional documentation please use:

```{r eval = FALSE}
if(!require(devtools)){install.package("devtools")}
devtools::install_github("khufkens/MODISTools", build_vignettes = TRUE)
library("MODISTools")
```

## Use

### Downloading MODIS time series

To extract a time series of modis data for a given location and its direct environment use the get_subset() function.

<details><summary>detailed parameter description (click to expand)</summary>
<p>

Parameter     | Description                      
------------- | ------------------------------ 	
product	      | a MODIS product
band	      | a MODIS product band (if NULL all bands are downloaded)
lat           | latitude of the site
lon           | longitude of the site
start      | start year of the time series (data start in 1980)
end        | end year of the time series (current year - 2 years, use force = TRUE to override)
internal      | logical, TRUE or FALSE, if true data is imported into R workspace otherwise it is downloaded into the current working directory
out_dir | path where to store the data when not used internally, defaults to tempdir()
km_lr | force "out of temporal range" downloads (integer)
km_ab | suppress the verbose output (integer)
site_name | a site identifier
site_id | a site_id for predefined locations (not required)

</p>
</details>

```{r eval = TRUE}
# load the library
library(MODISTools)

# download data
subset <- get_subset(product = "MOD11A2",
                    lat = 40,
                    lon = -110,
                    band = "LST_Day_1km",
                    start = "2004-01-01",
                    end = "2004-02-01",
                    km_lr = 1,
                    km_ab = 1,
                    site_name = "testsite",
                    internal = TRUE)
head(subset)
```

The output format is a nested list with the true data provided as a *tidy* data frame, as shown above. When witten to a csv, when the parameter 'internal' is set to FALSE, the same information is retained. Data can be read back into the same format with the included read_subset() function (see below).

Note that when a a region is defined using km_lr and km_ab multiple pixels might be returned. These are indexed using the 'pixel' column in the data frame containing the time series data. The remote sensing values are listed in the 'data' column. When no band is specified all bands of a given product are returned, be mindful of the fact that different bands might require different multipliers to represent their true values. 

To list all available products, bands for particular products and temporal coverage see function descriptions below.

### Batch downloading MODIS time series

Below an example is provided on how to batch download data for a data frame of given site names and locations (lat / lon).

```{r eval = TRUE}
# create data frame with a site_name, lat and lon column
# holding the respective names of sites and their location
df <- data.frame("site_name" = paste("test",1:2))
df$lat <- 40
df$lon <- -110
  
# test batch download
subsets <- batch_subset(df = df,
                     product = "MOD11A2",
                     band = "LST_Day_1km",
                     internal = TRUE,
                     start = "2004-01-01",
                     end = "2004-03-31",
                     out_dir = "~")

print(str(subsets))
```


### Listing products

To list all available products use the list_products() function.

```{r eval = TRUE}
products <- list_products()
head(products)
```


### Listing bands

To list all available bands for a given product use the list_bands() function.

```{r eval = TRUE}
bands <- list_bands(product = "MOD11A2")
head(bands)
```

### listing dates

To list all available dates (temporal coverage) for a given product and location use the list_dates() function.

```{r eval = TRUE}
dates <- list_dates(product = "MOD11A2", lat = 42, lon = -110)
head(dates)
```

### Reading and writing data from / to file

Data can be written to file using the write_subset() function which uses the following file format:

```
[site_name]_[product]_[start]_[end].csv
```

The data can be read back into the original nested structed using read_subset().

```{r eval = TRUE}
# write the above file to disk
write_subset(df = subset,
             out_dir = tempdir())

# read the data back in
subset_disk <- read_subset(paste0(tempdir(),
                  "/testsite_MOD11A2_2004-01-01_2004-02-01.csv"))

# compare original to read from disk
identical(subset, subset_disk)
```


