---
title: "MODISTools"
author: "Koen Hufkens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MODISTools functionality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# load the library
library(MODISTools)

```

The MODISTools package has as goal to facilitate the interface between R and the MODIS Land Product Subset API at the Oak Ridge National Laboratory DAAC. This programmatic interface to the ['MODIS Land Products Subsets' web services](https://modis.ornl.gov/data/modis_webservice.html) allows for easy downloads of 'MODIS' time series directly to your R workspace or your computer. Below an examples are provided on how to download a MODIS time series as well as list ancillary data.

### Listing products / bands / dates

In order to assess which products are available, which product bands are provided and which temporal range is covered one has to list these ancillary data. All these options can be queried using the `mt_*()` functions.

To list all available products use the mt_products() function.

```{r eval = TRUE}
products <- mt_products()
head(products)
```

To list all available bands for a given product use the mt_bands() function.

```{r eval = TRUE}
bands <- mt_bands(product = "MOD13Q1")
head(bands)
```

To list all available dates (temporal coverage) for a given product and location use the mt_dates() function.

```{r eval = TRUE}
dates <- mt_dates(product = "MOD13Q1", lat = 42, lon = -110)
head(dates)
```

### Downloading MODIS time series

Once you decide on which data to download using the above functions you can use these parameters to download a time series using the `mt_subset()` function. The below query downloads MOD11A2 based daytime lands surface temperatures. The location is named 'testsite' and only covers the first six months of 2004. The output is saved to a variable called subset in the R workspace (as defined by the parameter internal = TRUE, when set to FALSE the data is written to file).

```{r eval = TRUE}
# download data
subset <- mt_subset(product = "MOD13Q1",
                    lat = 42.534171,
                    lon = -72.179003,
                    band = "250m_16_days_NDVI",
                    start = "2004-01-01",
                    end = "2008-12-30",
                    km_lr = 0,
                    km_ab = 0,
                    site_name = "testsite",
                    internal = TRUE)
head(subset)

```

The output format is a *tidy* data frame, as shown above. When witten to a csv with the parameter `internal = FALSE` this will result in a flat file on disk.

Note that when a a region is defined using km_lr and km_ab multiple pixels might be returned. These are indexed using the `pixel` column in the data frame containing the time series data. The remote sensing values are listed in the `value` column. When no band is specified all bands of a given product are returned, be mindful of the fact that different bands might require different multipliers to represent their true values. 

When a large selection of locations is needed you might benefit from using the batch download function `mt_batch_subset()`, which provides a wrapper around the `mt_subset()` function in order to speed up large download batches. This function has a similar syntax to `mt_subset()` but requires a data frame defining site names (site_name) and locations (lat / lon) (or a comma delimited file with the same structure) to specify a list of download locations.

```{r eval = TRUE}
# create data frame with a site_name, lat and lon column
# holding the respective names of sites and their location
df <- data.frame("site_name" = paste("test",1:2), stringsAsFactors = FALSE)
df$lat <- 40
df$lon <- -110

# an example batch download data frame
print(df)

# test batch download
subsets <- mt_batch_subset(df = df,
                     product = "MOD11A2",
                     band = "LST_Day_1km",
                     internal = TRUE,
                     start = "2004-01-01",
                     end = "2004-02-28",
                     out_dir = "~")

head(subsets)
```


### Worked example using NDVI values

The above downloaded subset shows normalized difference vegetation index (NDVI), and [an indicator of plant health, density and or complexity](https://earthobservatory.nasa.gov/Features/MeasuringVegetation/measuring_vegetation_2.php) for the experimental [Harvard Forest](https://harvardforest.fas.harvard.edu/) site. When plotted the yearly pattern of rising and falling NDVI values shows vegetation growth in spring, and senescence or leaf drop in fall. Values lower than ~0.4 are contaminated by snow or clouds.

```{r fig.width = 7, fig.height=3}
# create a plot of the data - accounting for the multiplier (scale) component
date <- as.Date(subset$calendar_date)
temperature <- subset$value * as.double(subset$scale)
temperature[temperature == 0] <- NA

plot(date,
     temperature,
     xlab = "Date",
     ylab = "NDVI",
     ylim = c(0,1),
     type = "l")
```

### Conversion of corner coordinates

Corner coordinates of the pixel area extracted are provided, these can be used to calculate the coverage of the extracted area. Coordinates are provided in the original sinusoidal grid coordinates and first have to be transformed into latitude longitude (for convenience).

``` {r}
# convert the coordinates
subset <- subset %>%
  mutate(result = map2(xllcorner, yllcorner, sin_to_ll)) %>%
  unnest()
```

Together with meta-data regarding cell size, number of columns and rows the bounding box of the extracted data can be calculated.

``` {r eval = FALSE}
# convert the bounding box
subset <- subset %>% TODO
```
